X-Factors: External Influences of a Hockey Goaltender's Performance

### 2024-12-04 paper/paper.qmd
edited reference

To analyze the effect of external factors on goaltender performance, we use a log of 394,250 shots taken during the 2021-22 to 2024-25 NHL regular and post-seasons, provided by Moneypuck [@moneypuck]. Moneypuck is a publicly available source of hockey analytics that tracks advanced statistics, playoff odds, and event data. The shot data is updated daily from official NHL game reports, and is organized by season. For the current 2024-25 season, data is included up to November 26, 2024.

@@

Shots, defined as any player’s attempt to score by directing the puck at the net, are classified as an event type by the NHL. Events, along with their associated attributes and metadata, are officially recorded by Sportradar, the NHL's data partner, using AI-powered data tools [@sportradar]. Events are also counted and verified by official scorekeepers.

@@

- `last_game_GSAx`: GSAx from the previous game, representing goaltender performance in the previous game.  
- `last_5_avg_GSAx`: average GSAx from the 5 previous games, representing performance in the previous 5 games.

@@

summary_GSAx_segment <- data.frame(
  Statistic = names(summary_stats),
  Value = as.numeric(summary_stats)) %>%
  kable(format = "pipe", col.names = c("Summary Statistic", "GSAx_segment")) 
summary_GSAx_segment
```
From @tbl-1, the mean GSAx_segment is -0.01, close to the expected value of 0. This aligns with the idea that GSAx should average out to 0 when aggregated across many segments and goaltenders, as GSAx represents the difference between actual and expected goals. The median is slightly positive at 0.03, indicating that a slight majority of time segments may reflect slightly above-average performance. The standard deviation of 0.13 `GSAx_segment` indicates moderate variability in performance across time segments.

@@

summary_score <- data.frame(
  Statistic = names(summary_stats2),
  Value = as.numeric(summary_stats2)) %>%
  kable(format = "pipe", col.names = c("Summary Statistic", "Goalie Team Score Diff.")) 
summary_score
```
@tbl-5 shows this stat is symmetrical, as every non-tied segment has a winning and losing goalie. Both the median and mean are 0, representing tied segments. The median and mean is 0, representing segments where the game is tied. Due to its symmetry, we can better visualize the `goalieTeamScoreDifferential` distribution by taking the absolute value in @fig-2.

@@

From @tbl-7 and @fig-5, segments are roughly normally distributed with a mean of 1.3, meaning a goaltender saw ~1.3 shots in the 3 minutes prior to every shot in an average segment. Segments ranged from those with no shots in the preceding 3 minutes to segments where the goaltender saw an average of 9 shots in the last 3 minutes.

@@

summary_sofar <- data.frame(
  Statistic = names(summary_stats4),
  Value = as.numeric(summary_stats4)) %>%
  kable(format = "pipe", col.names = c("Summary Statistic", "GSAx so far")) 
summary_sofar
```
In @tbl-8, the median 0.12 and mean 0.1 suggest that goaltender performance prior to segments is generally close to expectations, with slight overperformance common. In @fig-6 the distribution is approximately normal and centered near 0; the range (-5.98 to 6.06) and standard deviation 1.01 indicate considerable variability in in-game performance.

@@

- `isHomeTeam`  
   Coefficient: `-0.003`  
   Implication: Playing at home is associated with a slight decrease in `GSAx_segment`, suggesting that goaltenders may slightly underperform at home.

- `period3rd Period`  
   Coefficient: `0.013`  
   Implication: Performance improves slightly in the 3rd period compared to the baseline 1st period.

- `periodOvertime`  
   Coefficient: `-0.091`  
   Implication: Goaltenders perform significantly worse during overtime compared to the baseline 1st period.

@@

- `goalieTeamScoreDifferential`  
   Coefficient: `0.011`  
   Implication: An increase in the score differential (goalie's team scores a goal) is associated with an increase in `GSAx_segment`, indicating improved goaltender performance when leading.

- `shotslast3min`  
   Coefficient: `0.009`  
   Implication: Facing more shots in the last three minutes is associated with a better `GSAx_segment`, suggesting that recent activity may keep the goaltender in a rhythm or alert.

- `shots_faced`  
   Coefficient: `-0.001`  
   Implication: Each additional shot faced is associated with a small decrease in `GSAx_segment`, suggesting that increased workload slightly reduces performance.

- `GSAx_so_far`  
   Coefficient: `0.022`  
   Implication: Better cumulative performance earlier in the game is associated with improved performance during the segment, indicating momentum or consistency.

- `shotslast3min * GSAx_so_far`  
   Coefficient: `0.002`  
   Implication: The interaction between recent activity (shots in the last three minutes) and cumulative performance (`GSAx_so_far`) suggests that recent activity amplifies the positive effects of prior performance.

@@

- Intercept: The baseline value itself is not particularly meaningful since most predictors will not realistically be zero.
- `isPlayoffGame`  
- `period2nd OT or later`  
- `period2nd Period`  
- `last_game_GSAx`  
- `last_5_avg_GSAx`  
- `isHomeTeam * isPlayoffGame` interaction
- `goalieTeamScoreDifferential * shotslast3min` interactoin

@@

### Goalie Team Score Differential
```{r}
#| echo: false
#| warning: false
#| label: tbl-scorediff
#| tbl-cap: GSAx_segment values by score differential

@@

### Shots faced in last 3 minutes
```{r}
#| echo: false
#| label: tbl-last3
#| tbl-cap: GSAx_segment values versus shots faced in last 3 minutes

@@

# Appendix {-}

@@

We calculate variables required for our analysis model: Goals Saved Above Expected (`GSAx`) for each shot is obtained by subtracting the expected goal value of the shot `xGoal` from boolean value `goal`. `xGoal` is modelled by Moneypuck based on many other shot variables [@moneypuck].

@@

### Checking relationship between predictors and residuals
```{r}
#| echo: false
#| warning: false
#| out-width: 50%

@@

### Normal Q-Q plot to check normality
```{r}
#| echo: false
#| out-width: 50%

ggplot(residuals_data, aes(sample = residuals)) +
  stat_qq(alpha = 0.5) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Q-Q Plot of Residuals",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```
<!-- ### Generalized Variance Inflation Factor -->

<!-- ```{r} -->
<!-- #| echo: false -->

<!-- vif <- vif(model, type = 'predictor') -->
<!-- vif %>% kable(format = "pipe", col.names = c("Predictor", "GVIF", "df", "GVIF^(1/(2*Df))", "Interacts With"))  -->
<!-- ``` -->

<!-- ### Validating with test data -->
<!-- ```{r} -->
<!-- #| echo: false -->

<!-- test_predictions <- predict(model, newdata = test_data) -->
<!-- test_residuals <- test_data$GSAx_segment - test_predictions -->
<!-- # Mean Squared Error (MSE) -->
<!-- mse <- mean(test_residuals^2) -->
<!-- print(paste("Mean Squared Error (MSE):", mse)) -->
<!-- # R-squared for the test set -->
<!-- ss_total <- sum((test_data$GSAx_segment - mean(test_data$GSAx_segment))^2) -->
<!-- ss_residual <- sum(test_residuals^2) -->
<!-- r_squared <- 1 - (ss_residual / ss_total) -->
<!-- print(paste("R-squared on test set:", r_squared)) -->
<!-- ``` -->

>>>>

To analyze the effect of external factors on goaltender performance, we use a log of 394,250 shots taken during the 2021-22 to 2024-25 NHL regular and post-seasons, provided by Moneypuck [(@moneypuck)]. Moneypuck is a publicly available source of hockey analytics that tracks advanced statistics, playoff odds, and event data. The shot data is updated daily from official NHL game reports, and is organized by season. For the current 2024-25 season, data is included up to November 26, 2024.

@@

Shots, defined as any player’s attempt to score by directing the puck at the net, are classified as an event type by the NHL. Events, along with their associated attributes and metadata, are officially recorded by Sportradar, the NHL's data partner, using AI-powered data tools [(@sportradar)]. Events are also counted and verified by official scorekeepers.

@@

-   `last_game_GSAx`: GSAx from the previous game, representing goaltender performance in the previous game.\
-   `last_5_avg_GSAx`: average GSAx from the 5 previous games, representing performance in the previous 5 games.

@@

summary_GSAx_segment <- data.frame(
  Statistic = names(summary_stats),
  Value = as.numeric(summary_stats)) %>%
  kable(format = "pipe", col.names = c("Summary Statistic", "GSAx_segment")) 
summary_GSAx_segment
```

From @tbl-1, the mean GSAx_segment is -0.01, close to the expected value of 0. This aligns with the idea that GSAx should average out to 0 when aggregated across many segments and goaltenders, as GSAx represents the difference between actual and expected goals. The median is slightly positive at 0.03, indicating that a slight majority of time segments may reflect slightly above-average performance. The standard deviation of 0.13 `GSAx_segment` indicates moderate variability in performance across time segments.

@@

summary_score <- data.frame(
  Statistic = names(summary_stats2),
  Value = as.numeric(summary_stats2)) %>%
  kable(format = "pipe", col.names = c("Summary Statistic", "Goalie Team Score Diff.")) 
summary_score
```

@tbl-5 shows this stat is symmetrical, as every non-tied segment has a winning and losing goalie. Both the median and mean are 0, representing tied segments. The median and mean is 0, representing segments where the game is tied. Due to its symmetry, we can better visualize the `goalieTeamScoreDifferential` distribution by taking the absolute value in @fig-2.

@@

From @tbl-7 and @fig-5, segments are roughly normally distributed with a mean of 1.3, meaning a goaltender saw \~1.3 shots in the 3 minutes prior to every shot in an average segment. Segments ranged from those with no shots in the preceding 3 minutes to segments where the goaltender saw an average of 9 shots in the last 3 minutes.

@@

summary_sofar <- data.frame(
  Statistic = names(summary_stats4),
  Value = as.numeric(summary_stats4)) %>%
  kable(format = "pipe", col.names = c("Summary Statistic", "GSAx so far")) 
summary_sofar
```

In @tbl-8, the median 0.12 and mean 0.1 suggest that goaltender performance prior to segments is generally close to expectations, with slight overperformance common. In @fig-6 the distribution is approximately normal and centered near 0; the range (-5.98 to 6.06) and standard deviation 1.01 indicate considerable variability in in-game performance.

@@

-   `isHomeTeam`\
    Coefficient: `-0.003`\
    Implication: Playing at home is associated with a slight decrease in `GSAx_segment`, suggesting that goaltenders may slightly underperform at home.

-   `period3rd Period`\
    Coefficient: `0.013`\
    Implication: Performance improves slightly in the 3rd period compared to the baseline 1st period.

-   `periodOvertime`\
    Coefficient: `-0.091`\
    Implication: Goaltenders perform significantly worse during overtime compared to the baseline 1st period.

@@

-   `goalieTeamScoreDifferential`\
    Coefficient: `0.011`\
    Implication: An increase in the score differential (goalie's team scores a goal) is associated with an increase in `GSAx_segment`, indicating improved goaltender performance when leading.

-   `shotslast3min`\
    Coefficient: `0.009`\
    Implication: Facing more shots in the last three minutes is associated with a better `GSAx_segment`, suggesting that recent activity may keep the goaltender in a rhythm or alert.

-   `shots_faced`\
    Coefficient: `-0.001`\
    Implication: Each additional shot faced is associated with a small decrease in `GSAx_segment`, suggesting that increased workload slightly reduces performance.

-   `GSAx_so_far`\
    Coefficient: `0.022`\
    Implication: Better cumulative performance earlier in the game is associated with improved performance during the segment, indicating momentum or consistency.

-   `shotslast3min * GSAx_so_far`\
    Coefficient: `0.002`\
    Implication: The interaction between recent activity (shots in the last three minutes) and cumulative performance (`GSAx_so_far`) suggests that recent activity amplifies the positive effects of prior performance.

@@

-   Intercept: The baseline value itself is not particularly meaningful since most predictors will not realistically be zero.
-   `isPlayoffGame`\
-   `period2nd OT or later`\
-   `period2nd Period`\
-   `last_game_GSAx`\
-   `last_5_avg_GSAx`\
-   `isHomeTeam * isPlayoffGame` interaction
-   `goalieTeamScoreDifferential * shotslast3min` interactoin

@@

### Goalie Team Score Differential

```{r}
#| echo: false
#| warning: false
#| label: tbl-scorediff
#| tbl-cap: GSAx_segment values by score differential

@@

### Shots faced in last 3 minutes

```{r}
#| echo: false
#| label: tbl-last3
#| tbl-cap: GSAx_segment values versus shots faced in last 3 minutes

@@

# Appendix {.unnumbered}

@@

We calculate variables required for our analysis model: Goals Saved Above Expected (`GSAx`) for each shot is obtained by subtracting the expected goal value of the shot `xGoal` from boolean value `goal`. `xGoal` is modelled by Moneypuck based on many other shot variables [(@moneypuck)].

@@

### Checking relationship between predictors and residuals

```{r}
#| echo: false
#| warning: false
#| out-width: 50%

@@

### Normal Q-Q plot to check normality

ggplot(residuals_data, aes(sample = residuals)) +
  stat_qq(alpha = 0.5) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Q-Q Plot of Residuals",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

<!-- ### Generalized Variance Inflation Factor -->

<!-- ```{r} -->

<!-- #| echo: false -->

<!-- vif <- vif(model, type = 'predictor') -->

<!-- vif %>% kable(format = "pipe", col.names = c("Predictor", "GVIF", "df", "GVIF^(1/(2*Df))", "Interacts With"))  -->

<!-- ``` -->

<!-- ### Validating with test data -->

<!-- ```{r} -->

<!-- #| echo: false -->

<!-- test_predictions <- predict(model, newdata = test_data) -->

<!-- test_residuals <- test_data$GSAx_segment - test_predictions -->

<!-- # Mean Squared Error (MSE) -->

<!-- mse <- mean(test_residuals^2) -->

<!-- print(paste("Mean Squared Error (MSE):", mse)) -->

<!-- # R-squared for the test set -->

<!-- ss_total <- sum((test_data$GSAx_segment - mean(test_data$GSAx_segment))^2) -->

<!-- ss_residual <- sum(test_residuals^2) -->

<!-- r_squared <- 1 - (ss_residual / ss_total) -->

<!-- print(paste("R-squared on test set:", r_squared)) -->

<!-- ``` -->
